Computing Direct Sum Decompositions

### 2025-04-16 paper/main.tex
macro changes

The purpose of this paper is to describe and prove correctness of a practical algorithm for computing indecomposable summands of finitely generated modules over a finitely generated local or graded $k$-algebra, for $k$ a field of positive characteristic. We also describe applications of our algorithm for computations in characteristic 0. Further, our algorithm takes advantage of multigradings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in $\P^n$).

@@

\begin{itemize}
\item For a field $k$, we will write $\overline k$ for a choice of algebraic closure of $k$. In particular, $\overline{\mathbb F_p}$ is the algebraic closure of the finite field $\mathbb F_p$.
\item For an algebraic variety $X$ and a point $x\in X$, we write $k(x)=\O_{X,x}/m_x$ for the residue field of $X$ at $x$.
\item We write $k(X)$ for the function field of $X$.
\item If $X$ is a projective variety, and $\O_X(1)$ a fixed very ample line bundle on $X$, then for any coherent $\O_X$-module $\cF$ we write 
$\Gamma_* \cF = \bigoplus_{n\geq 0} H(X,\cF(n))$. This is a module over the graded ring $S=\bigoplus_{n\geq 0} H^0(X,\O_X(n))$.
\item Conversely, if $S$ is a graded ring defining a projective variety $X$, and $M$ is a finitely generated $S$-module, we write $\wtilde M$ for the coherent sheaf on $X$ associated to $M$.
\item For a graded or local ring $(R,m)$ and an $R$-module $M$, we write $\mu(M):=\dim_{R/m} M/mM$ for the minimal number of generators of $M$ as an $R$-module.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The main algorithms} 
Here, we describe the main algorithms for finding the indecomposable summands of a finitely generated $R$-module $M$, when $R$ is either a graded ring or a local ring with residue field contained in $\overline{\mathbb F_p}$.

\subsection{The graded algorithm}\label{sec:main-alg-new}
Let $R$ be a $\N$-graded ring, with $R_0=k$ an algebraically closed field.
%Assume furthermore that the degree support of $R$ is strongly convex, i.e., that 
%%the effective cone
%$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\Z^n$.
Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree-0 endomorphism.

%Let $P\subset \Z^n$ be the degrees of a set of minimal generators of $M$; 
%this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
%Because $P$ is finite, we can order the elements of $P$ such that 
%$\deg m_1\not\leq \deg m_2\not \leq \dots\not\leq \deg m_r$.

@@

\begin{rem}[how to do it in multigraded case]
Let $R$ be an $\Z^n$-graded ring, with $R_0=k$ an algebraically closed field.
Assume furthermore that the degree support of $R$ is strongly convex, i.e., that 
the effective cone
$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\Z^n$.
Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree-0 endomorphism.

Let $P\subset \Z^n$ be the degrees of a set of minimal generators of $M$; 
this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
Because $P$ is finite, we can order the elements of $P$ such that 
$\deg m_1\not\leq \deg m_2\not \leq \dots\not\leq \deg m_r$.

@@

Throughout this section, $(R,m)$ will be either local with maximal ideal $m$, or a (multi)graded ring with $R_0 = k \subset \overline{\mathbb F_p}$ a field of positive characteristic and homogeneous maximal ideal $m$.

@@

\begin{lem}\label{lem:jordan}
  Let $k$ be a finite field of characteristic $p$, and let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$. If $\l$ is an eigenvalue of $A$, then some power of $A-\l I$ is idempotent.
Furthermore, if $\l$ is not the only eigenvalue of $A$, then a  nonzero power of $A-\l I$ is idempotent.
\end{lem}
\begin{proof}
  Since all eigenvalues of $A$ are contained in $k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block having the form
  \[ r_i\Biggl\{
  \underbrace{\begin{pmatrix}
      \l_i & 1 & 0 & \dots & 0 \\
      0 & \l_i & 1  & \dots & 0 \\
      \vdots & & & & \vdots \\
      0 & 0 & 0 & \dots & \l_i
  \end{pmatrix}}_{r_i} \]
  with each $\l_i$ an eigenvalue of $A$.
  In this basis, $A-\l I$ will be block-diagonal with blocks of form
  \[ \begin{pmatrix}
    \l_i-\l & 1 & 0 & \dots & 0 \\
    0 & \l_i-\l & 1 & \dots & 0 \\
    \vdots & & & & \vdots \\
    0 & 0 & 0 & \dots & \l_i-\l
  \end{pmatrix}. \]
  Set $\mu_i=\l_i-\l$. Then for any $l\geq 1$, the $l$-th power $(A-\l I)^l$ is block-diagonal with blocks of the form
  \[ \begin{pmatrix}
    \mu_i^l & \binom{l}1\mu_i^{l-1} & \binom{l}2\mu_i^{l-2} & \dots & \binom{l}{r_i}\mu_i^{l-r_i} \\
    0 & \mu_i^l & \binom{l}1\mu_i^{l-1} & \dots & \binom{l}{r_i-1}\mu_i^{l-r_i+1} \\
    \vdots & & & & \vdots \\
    0 & 0 & 0 & \dots & \mu_i^l
  \end{pmatrix}. \]
  If we choose $l > r_i$ and $l$ a power of $p$, then all non-diagonal terms will vanish, so all blocks will have the form
  \[ \begin{pmatrix}
    \mu_i^l & 0 & 0 & \dots & 0 \\
    0 & \mu_i^l & 0 & \dots & 0 \\
    \vdots & & & & \vdots \\
    0 & 0 & 0 & \dots & \mu_i^l
  \end{pmatrix}. \]
  Finally, since $k=\mathbb F_{p^e}$ for some $e$, if we choose $l$ to be divisible also by $p^e-1$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1.$$
  Thus, $(A-\l I)^l$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.
\end{proof}

@@

\begin{rem}[Characteristic zero]
If $R$ is a local ring over a field $k\subset \C$ of characteristic 0, and $M$ a finitely generated $R$-module, it would be useful to have an algorithm to find a direct sum decomposition of $M$.
At the moment, we do not have such an algorithm outside the graded case.
However, we note that the algorithm above can be used to test indecomposability of $M$ via reduction modulo $p$, as follows:
One can choose a finitely generated $\Z$-algebra $A$ and an $A$-algebra $R_A$ such that $R_A\otimes _A k=R$, and likewise an $R_A$-module $M_A$ such that $M_A\otimes _A k = R$ and such that $M_A$ is flat over $A$. If $n$ is a maximal ideal of $A$, one can check that $A/n \cong \mathbb F_{p^e}$ for some prime $p$ and $e$.
The key point is that if $M$ is decomposable, then $M_A$ also can be taken to be decomposable. Thus, the various reductions  $M\otimes A/n$ will also be decomposable for all $n$.
If our algorithm does not detect an indecomposable summand of $M\otimes A/n$ for an appropriate choice of $n$, then, the original module $M$ must have been indecomposable.
However, if $M$ is \emph{decomposable}, we do not have a way of patching the decompositions of various $M_A\otimes_A A/n$ into a decomposition of $M$.

@@

Let $X\subset \P^n$ be a projective variety, with ample line bundle $\O_X(1)$.
Throughout, let $E$ be a vector bundle (i.e., a locally free coherent sheaf) on $X$.

@@

%\begin{proof}
%This requires only that $E$ is a coherent sheaf:
%% since $\End_{\O_X}(E)$ is finite-dimensional, $f$ satisfies some minimal monic polynomial $\phi(x)$; since $f$ is injective, the constant term of $\phi(x)$ is nonzero.
%We claim that $f\otimes k(x)$ is injective for any $k(x)$. This then implies that $f\otimes k(x)$ is surjective for every $x\in X$ and thus $f$ is surjective.
%To see this, note that $f$ satisfies some minimal-degree monic polynomial, since $\End_{\O_X}(E)$ is finite-dimensional over $k$, and that this monic polynomial has nonzero constant term, since $f$ is injective. Now, $f\otimes k(x)$ satisfies this same polynomial, so must be injective.
%\end{proof}

\begin{dfn}
For $f\in \End_{\O_X}(E)$, taking top exterior powers yields the map
$$ \bigwedge^{\rank E} f: \det E  \to \det E, $$
where $\det E := \bigwedge^{\rank E} E$.
Since $\End_{\O_X}(\det E) = H^0(\O_X)=k$, $\bigwedge^{\rank E} f$ is multiplication by $\l\in k$; we write $\det f$ for this scalar $\l$.
\end{dfn}

\begin{lem}
Let $x \in X$ be any point (not necessarily closed), with residue field $k(x) = \O_{X,x}/m_x$. Then $\det(f\otimes k(x)) =\det f$.
%In other words, if we restrict $f$ to the fiber $E_x$, the determinant is the same as $\det f$.
\end{lem}

@@

\begin{dfn}
$\l \in k$ is an \defn{eigenvalue} of $f \in \End_{\O_X}(E)$ if $\det(f-\l \id_E)= 0$. In other words, the eigenvalues of $f$ are the zeroes of the univariate polynomial $\det(f-t\id _E)$ with coefficients in $k$.
\end{dfn}

@@

\begin{prop}
If $X$ is a variety over a field $k\subset \overline{\mathbb F_{p}}$,
$E$ a vector bundle on $X$, and $f:E\to E$ has two distinct eigenvalues $\l_1,\l_2$, then $E$ is a nontrivial direct sum, with one summand given by $(f-\l_1)^{p^e(p^e-1)}$ for some $e\geq 1$.
\end{prop}

@@

\begin{exa}[Frobenius pushforward on the projective space $\P^n$]
  Let $S = k[x_0,\dots,x_n]$ be a polynomial ring with $\char k = p$ and $\deg x_i = 1$ and consider the Frobenius endomorphism
  \[ F\colon S\to S \quad \text{given by} \quad f \to f^p. \]
  Hartshorne \cite{Hartshorne1970} proved that for any line bundle $L\in\Pic\P^n$, the Frobenius pushforward $F_*L$ splits as a sum of line bundles. While the following calculations are straightforward to do by hand, they are immediately calculated via our algorithm:
  \begin{align*}
    \text{When }p=3, n=2: \\
    F_*\O_{\P^2} &= \O \oplus \O(-1)^7 \oplus \O(-2). \\
    \text{When } p=2, n=5: \\
    F_*\O_{\P^5} &= \O \oplus \O(-1)^{15} \oplus \O(-2)^{15} \oplus \O(-3), \\
    F_*^2\O_{\P^5} &= \O \oplus \O(-1)^{120} \oplus \O(-2)^{546} \oplus \O(-3)^{336} \oplus \O(-4)^{21}.
  \end{align*}
\end{exa}

\begin{exa}[Frobenius pushforward on toric varieties]
  Let $X$ be a smooth toric variety and consider its Cox ring
  \[ S = \bigoplus_{[D]\in\Pic{X}} \; \Gamma(X, \O(D)). \]
  Similar to the case of the projective space, B{\o}gvad and Thomsen \cite{Bogvad98,Thomsen00} showed that $F_*L$ totally splits as a direct sum of line bundles for any line bundle $L\in\Pic X$.

  As an example, consider the third Hirzebruch surface $X=\P(\O_{\P^1}\oplus \O_{\P^1}(3))$ over a field of characteristic 3. We have, for example, that
  \begin{align*}
    F_*\O_X      &= \O_X   \oplus \O_X(-1,0)^2 \oplus \O_X(0,-1)^2 \oplus \O_X(1,-1)^3 \oplus \O_X(2,-1), \\
    F_*\O_X(1,1) &= \O_X^3 \oplus \O_X(-1,0)   \oplus \O_X(1,-1)   \oplus \O_X(1, 0)^2 \oplus \O_X(2,-1)^2.
  \end{align*}
  In fact, Achinger \cite{Achinger15} showed that the total splitting of $F_*L$ for every line bundle $L$ characterizes smooth projective toric varieties.
  % The toric Frobenius is simply multiplication by $p$ on the torus.
  % The $p$ need not be prime, nor even match the characteristic!
\end{exa}

\begin{exa}[Frobenius pushforward on elliptic curves]\label{algclos}
  Consider the elliptic curve
  \[ X = \Proj \mathbb F_7[x,y,z]/(x^3+y^3+z^3). \]
  %%Weierstrauss form is y^2=x^3-432
  This is an ordinary elliptic curve, hence $F$-split; thus $\O_X$ is a summand of $F_* \O_X$. Over the algebraic closure of $\mathbb F_7$, $F_*\O_X$ will decompose as $\bigoplus_{p=1}^7 \O_X(p_i)$, where $p_1,\dots,p_7$ are the 7-torsion points of $X$.

  However, over $\mathbb F_7$, our algorithm calculates that $F_*\O_X$ decomposes only as
  $$ F_* \O_X =\O_X \oplus M_1\oplus M_2\oplus M_3, $$
  with $M_i$ indecomposable (over $\mathbb F_7$) of rank 2.

  If one extends the ground field to $\mathbb F_{49}$, however, our algorithm calculates the full decomposition
  $$ F_* \O_X=\bigoplus_{p=1}^7 \O_X(p_i). $$
  This reflects the fact that the 7-torsion points $p_i$ of $X$, and thus the sheaves $\O_X(p_i)$, are not defined over $\mathbb F_7$, but rather are defined over $\mathbb F_{49}$.
\end{exa}

\begin{exa}[Frobenius pushforward on Grassmannians]
  % https://arxiv.org/abs/1901.10956
  Consider the Grassmannian $X = \Gr(2,4)$. We may work over the Cox ring $S$,
  %\[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \O(D)). \]
  which in this case coincides with the coordinate ring
  \[ S = \frac{k[p_{0,1},p_{0,2},p_{0,3},p_{1,2},p_{1,3},p_{2,3}]}{p_{1,2}p_{0,3}-p_{0,2}p_{1,3}+p_{0,1}p_{2,3}}. \]
  Then in characteristic $p=3$ we have:
  \[ F_*\O_X = \O \oplus \O(-1)^{44} \oplus \O(-2)^{20} \oplus A^4 \oplus B^4, \]
  where $A$ and $B$ are rank-2 indecomposable bundles (c.f.~\cite{RSVdB22}).
\end{exa}

@@

  For instance, consider $X = \Bl_4\P^2$, the blowup of $\P^2$ at 4 general points. We will be working over the $\Z^5$-graded Cox ring
  \[ S = k[x_1,\dots,x_{10}]/\text{(five quadric Pl\"ucker relations)} \]
  with degrees
    \[
    \left(\!\begin{array}{rrrrrrrrrr}
      0&0&0&0&1&1&1&1&1&1 \\
      1&0&0&0&-1&-1&-1&0&0&0 \\
      0&1&0&0&-1&0&0&-1&-1&0 \\
      0&0&1&0&0&-1&0&-1&0&-1 \\
      0&0&0&1&0&0&-1&0&-1&-1
    \end{array}\!\right).
    \]
    Then in characteristic 2 we have:
    \begin{align*}
      F_*^2\O_X = {\O_{X}^{1}}
      &\oplus {\O_{X}^{2}\ \left(-2,\,1,\,1,\,1,\,1\right)} \oplus {\O_{X}^{2}\ \left(-1,\,0,\,0,\,0,\,1\right)} \\
      &\oplus {\O_{X}^{2}\ \left(-1,\,0,\,0,\,1,\,0\right)} \oplus {\O_{X}^{2}\ \left(-1,\,0,\,1,\,0,\,0\right)} \\
      &\oplus {\O_{X}^{2}\ \left(-1,\,1,\,0,\,0,\,0\right)} \oplus B \oplus G,
    \end{align*}
    where $B, G$ are rank-3 and rank-2 indecomposable modules, as calculated in \cite{Hara15}.
\end{exa}

\begin{exa}[Frobenius pushforward on cubic surfaces]
  Let $X$ be a smooth cubic surface. Aside from a single exception in characteristic 0, $X$ will be globally $F$-split, so that any $F^e_*\O_X $ admits $\O_X$ as a direct summand.
  The other summands of Frobenius pushforwards of $\O_X$ have yet to be studied, and in particular it is not known whether such rings should have the finite $F$-representation type property.

  The use of our algorithm to compute examples in small $p$ and $e$ suggest the following behavior:
  $$ F_* \O_X = \O_X\oplus M, $$
  with $M$ indecomposable, and furthermore $F_*^e M$ remains indecomposable for all $e\geq 0$. In other words, the indecomposable decomposition of $F^e_* \O_X$ is
  $$ F_*^e \O_X \cong \O_X\oplus M\oplus F_* M\oplus\dots\oplus F_*^{e-1}M. $$
  In particular, this suggests $\O_X$ will fail to have the finite $F$-representation type property.
  In fact, we believe a similar description holds true for quartic del Pezzos.
\end{exa}

>>>>

The purpose of this paper is to describe and prove correctness of a practical algorithm for computing indecomposable summands of finitely generated modules over a finitely generated local or graded $k$-algebra, for $k$ a field of positive characteristic. We also describe applications of our algorithm for computations in characteristic 0. Further, our algorithm takes advantage of multigradings, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in $\PP^n$).

@@

\begin{itemize}
\item For a field $k$, we will write $\overline k$ for a choice of algebraic closure of $k$. In particular, $\overline{\FF_p}$ is the algebraic closure of the finite field $\FF_p$.
\item For an algebraic variety $X$ and a point $x\in X$, we write $k(x)=\OO_{X,x}/m_x$ for the residue field of $X$ at $x$.
\item We write $k(X)$ for the function field of $X$.
\item If $X$ is a projective variety, and $\OO_X(1)$ a fixed very ample line bundle on $X$, then for any coherent $\OO_X$-module $\cF$ we write
$\Gamma_* \cF = \bigoplus_{n\geq 0} H(X,\cF(n))$. This is a module over the graded ring $S=\bigoplus_{n\geq 0} H^0(X,\OO_X(n))$.
\item Conversely, if $S$ is a graded ring defining a projective variety $X$, and $M$ is a finitely generated $S$-module, we write $\wtilde M$ for the coherent sheaf on $X$ associated to $M$.
\item For a graded or local ring $(R,m)$ and an $R$-module $M$, we write $\mu(M):=\dim_{R/m} M/mM$ for the minimal number of generators of $M$ as an $R$-module.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The main algorithms}
Here, we describe the main algorithms for finding the indecomposable summands of a finitely generated $R$-module $M$, when $R$ is either a graded ring or a local ring with residue field contained in $\overline{\FF_p}$.

\subsection{The graded algorithm}\label{sec:main-alg-new}
Let $R$ be a $\N$-graded ring, with $R_0=k$ an algebraically closed field.
%Assume furthermore that the degree support of $R$ is strongly convex, i.e., that
%%the effective cone
%$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree-0 endomorphism.

%Let $P\subset \ZZ^n$ be the degrees of a set of minimal generators of $M$;
%this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
%Because $P$ is finite, we can order the elements of $P$ such that
%$\deg m_1\not\leq \deg m_2\not \leq \dots\not\leq \deg m_r$.

@@

\begin{rem}[how to do it in multigraded case]
Let $R$ be an $\ZZ^n$-graded ring, with $R_0=k$ an algebraically closed field.
Assume furthermore that the degree support of $R$ is strongly convex, i.e., that
the effective cone
$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
Let $M$ be a finitely generated graded $R$-module, and let $\phi \in \End M$ be a degree-0 endomorphism.

Let $P\subset \ZZ^n$ be the degrees of a set of minimal generators of $M$;
this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
Because $P$ is finite, we can order the elements of $P$ such that
$\deg m_1\not\leq \deg m_2\not \leq \dots\not\leq \deg m_r$.

@@

Throughout this section, $(R,m)$ will be either local with maximal ideal $m$, or a (multi)graded ring with $R_0 = k \subset \overline{\FF_p}$ a field of positive characteristic and homogeneous maximal ideal $m$.

@@

\begin{lem}\label{lem:jordan}
  Let $k$ be a finite field of characteristic $p$, and let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$. If $\l$ is an eigenvalue of $A$, then some power of $A-\l I$ is idempotent.
Furthermore, if $\l$ is not the only eigenvalue of $A$, then a  nonzero power of $A-\l I$ is idempotent.
\end{lem}
\begin{proof}
  Since all eigenvalues of $A$ are contained in $k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block having the form
  \[ r_i\Biggl\{
  \underbrace{\begin{pmatrix}
      \l_i & 1 & 0 & \dots & 0 \\
      0 & \l_i & 1  & \dots & 0 \\
      \vdots & & & & \vdots \\
      0 & 0 & 0 & \dots & \l_i
  \end{pmatrix}}_{r_i} \]
  with each $\l_i$ an eigenvalue of $A$.
  In this basis, $A-\l I$ will be block-diagonal with blocks of form
  \[ \begin{pmatrix}
    \l_i-\l & 1 & 0 & \dots & 0 \\
    0 & \l_i-\l & 1 & \dots & 0 \\
    \vdots & & & & \vdots \\
    0 & 0 & 0 & \dots & \l_i-\l
  \end{pmatrix}. \]
  Set $\mu_i=\l_i-\l$. Then for any $l\geq 1$, the $l$-th power $(A-\l I)^l$ is block-diagonal with blocks of the form
  \[ \begin{pmatrix}
    \mu_i^l & \binom{l}1\mu_i^{l-1} & \binom{l}2\mu_i^{l-2} & \dots & \binom{l}{r_i}\mu_i^{l-r_i} \\
    0 & \mu_i^l & \binom{l}1\mu_i^{l-1} & \dots & \binom{l}{r_i-1}\mu_i^{l-r_i+1} \\
    \vdots & & & & \vdots \\
    0 & 0 & 0 & \dots & \mu_i^l
  \end{pmatrix}. \]
  If we choose $l > r_i$ and $l$ a power of $p$, then all non-diagonal terms will vanish, so all blocks will have the form
  \[ \begin{pmatrix}
    \mu_i^l & 0 & 0 & \dots & 0 \\
    0 & \mu_i^l & 0 & \dots & 0 \\
    \vdots & & & & \vdots \\
    0 & 0 & 0 & \dots & \mu_i^l
  \end{pmatrix}. \]
  Finally, since $k=\FF_{p^e}$ for some $e$, if we choose $l$ to be divisible also by $p^e-1$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1.$$
  Thus, $(A-\l I)^l$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.
\end{proof}

@@

\begin{rem}[Characteristic zero]
If $R$ is a local ring over a field $k\subset \C$ of characteristic 0, and $M$ a finitely generated $R$-module, it would be useful to have an algorithm to find a direct sum decomposition of $M$.
At the moment, we do not have such an algorithm outside the graded case.
However, we note that the algorithm above can be used to test indecomposability of $M$ via reduction modulo $p$, as follows:
One can choose a finitely generated $\Z$-algebra $A$ and an $A$-algebra $R_A$ such that $R_A\otimes _A k=R$, and likewise an $R_A$-module $M_A$ such that $M_A\otimes _A k = R$ and such that $M_A$ is flat over $A$. If $n$ is a maximal ideal of $A$, one can check that $A/n \cong \FF_{p^e}$ for some prime $p$ and $e$.
The key point is that if $M$ is decomposable, then $M_A$ also can be taken to be decomposable. Thus, the various reductions  $M\otimes A/n$ will also be decomposable for all $n$.
If our algorithm does not detect an indecomposable summand of $M\otimes A/n$ for an appropriate choice of $n$, then, the original module $M$ must have been indecomposable.
However, if $M$ is \emph{decomposable}, we do not have a way of patching the decompositions of various $M_A\otimes_A A/n$ into a decomposition of $M$.

@@

Let $X\subset \P^n$ be a projective variety, with ample line bundle $\OO_X(1)$.
Throughout, let $E$ be a vector bundle (i.e., a locally free coherent sheaf) on $X$.

@@

%\begin{proof}
%This requires only that $E$ is a coherent sheaf:
%% since $\End_{\OO_X}(E)$ is finite-dimensional, $f$ satisfies some minimal monic polynomial $\phi(x)$; since $f$ is injective, the constant term of $\phi(x)$ is nonzero.
%We claim that $f\otimes k(x)$ is injective for any $k(x)$. This then implies that $f\otimes k(x)$ is surjective for every $x\in X$ and thus $f$ is surjective.
%To see this, note that $f$ satisfies some minimal-degree monic polynomial, since $\End_{\OO_X}(E)$ is finite-dimensional over $k$, and that this monic polynomial has nonzero constant term, since $f$ is injective. Now, $f\otimes k(x)$ satisfies this same polynomial, so must be injective.
%\end{proof}

\begin{dfn}
For $f\in \End_{\OO_X}(E)$, taking top exterior powers yields the map
$$ \bigwedge^{\rank E} f: \det E  \to \det E, $$
where $\det E := \bigwedge^{\rank E} E$.
Since $\End_{\OO_X}(\det E) = H^0(\OO_X)=k$, $\bigwedge^{\rank E} f$ is multiplication by $\l\in k$; we write $\det f$ for this scalar $\l$.
\end{dfn}

\begin{lem}
Let $x \in X$ be any point (not necessarily closed), with residue field $k(x) = \OO_{X,x}/m_x$. Then $\det(f\otimes k(x)) =\det f$.
%In other words, if we restrict $f$ to the fiber $E_x$, the determinant is the same as $\det f$.
\end{lem}

@@

\begin{dfn}
$\l \in k$ is an \defn{eigenvalue} of $f \in \End_{\OO_X}(E)$ if $\det(f-\l \id_E)= 0$. In other words, the eigenvalues of $f$ are the zeroes of the univariate polynomial $\det(f-t\id _E)$ with coefficients in $k$.
\end{dfn}

@@

\begin{prop}
If $X$ is a variety over a field $k\subset \overline{\FF_{p}}$,
$E$ a vector bundle on $X$, and $f:E\to E$ has two distinct eigenvalues $\l_1,\l_2$, then $E$ is a nontrivial direct sum, with one summand given by $(f-\l_1)^{p^e(p^e-1)}$ for some $e\geq 1$.
\end{prop}

@@

\begin{exa}[Frobenius pushforward on the projective space $\P^n$]
  Let $S = k[x_0,\dots,x_n]$ be a polynomial ring with $\char k = p$ and $\deg x_i = 1$ and consider the Frobenius endomorphism
  \[ F\colon S\to S \quad \text{given by} \quad f \to f^p. \]
  Hartshorne \cite{Hartshorne1970} proved that for any line bundle $L\in\Pic\P^n$, the Frobenius pushforward $F_*L$ splits as a sum of line bundles. While the following calculations are straightforward to do by hand, they are immediately calculated via our algorithm:
  \begin{align*}
    \text{When }p=3, n=2: \\
    F_*\OO_{\P^2} &= \OO \oplus \OO(-1)^7 \oplus \OO(-2). \\
    \text{When } p=2, n=5: \\
    F_*\OO_{\P^5} &= \OO \oplus \OO(-1)^{15} \oplus \OO(-2)^{15} \oplus \OO(-3), \\
    F_*^2\OO_{\P^5} &= \OO \oplus \OO(-1)^{120} \oplus \OO(-2)^{546} \oplus \OO(-3)^{336} \oplus \OO(-4)^{21}.
  \end{align*}
\end{exa}

\begin{exa}[Frobenius pushforward on toric varieties]
  Let $X$ be a smooth toric variety and consider its Cox ring
  \[ S = \bigoplus_{[D]\in\Pic{X}} \; \Gamma(X, \OO(D)). \]
  Similar to the case of the projective space, B{\o}gvad and Thomsen \cite{Bogvad98,Thomsen00} showed that $F_*L$ totally splits as a direct sum of line bundles for any line bundle $L\in\Pic X$.

  As an example, consider the third Hirzebruch surface $X=\P(\OO_{\P^1}\oplus \OO_{\P^1}(3))$ over a field of characteristic 3. We have, for example, that
  \begin{align*}
    F_*\OO_X      &= \OO_X   \oplus \OO_X(-1,0)^2 \oplus \OO_X(0,-1)^2 \oplus \OO_X(1,-1)^3 \oplus \OO_X(2,-1), \\
    F_*\OO_X(1,1) &= \OO_X^3 \oplus \OO_X(-1,0)   \oplus \OO_X(1,-1)   \oplus \OO_X(1, 0)^2 \oplus \OO_X(2,-1)^2.
  \end{align*}
  In fact, Achinger \cite{Achinger15} showed that the total splitting of $F_*L$ for every line bundle $L$ characterizes smooth projective toric varieties.
  % The toric Frobenius is simply multiplication by $p$ on the torus.
  % The $p$ need not be prime, nor even match the characteristic!
\end{exa}

\begin{exa}[Frobenius pushforward on elliptic curves]\label{algclos}
  Consider the elliptic curve
  \[ X = \Proj \FF_7[x,y,z]/(x^3+y^3+z^3). \]
  %%Weierstrauss form is y^2=x^3-432
  This is an ordinary elliptic curve, hence $F$-split; thus $\OO_X$ is a summand of $F_* \OO_X$. Over the algebraic closure of $\FF_7$, $F_*\OO_X$ will decompose as $\bigoplus_{p=1}^7 \OO_X(p_i)$, where $p_1,\dots,p_7$ are the 7-torsion points of $X$.

  However, over $\FF_7$, our algorithm calculates that $F_*\OO_X$ decomposes only as
  $$ F_* \OO_X =\OO_X \oplus M_1\oplus M_2\oplus M_3, $$
  with $M_i$ indecomposable (over $\FF_7$) of rank 2.

  If one extends the ground field to $\FF_{49}$, however, our algorithm calculates the full decomposition
  $$ F_* \OO_X=\bigoplus_{p=1}^7 \OO_X(p_i). $$
  This reflects the fact that the 7-torsion points $p_i$ of $X$, and thus the sheaves $\OO_X(p_i)$, are not defined over $\FF_7$, but rather are defined over $\FF_{49}$.
\end{exa}

\begin{exa}[Frobenius pushforward on Grassmannians]
  % https://arxiv.org/abs/1901.10956
  Consider the Grassmannian $X = \Gr(2,4)$. We may work over the Cox ring $S$,
  %\[ S = \bigoplus\!{}_{[D]\in\Pic{X}} \; \Gamma(X, \OO(D)). \]
  which in this case coincides with the coordinate ring
  \[ S = \frac{k[p_{0,1},p_{0,2},p_{0,3},p_{1,2},p_{1,3},p_{2,3}]}{p_{1,2}p_{0,3}-p_{0,2}p_{1,3}+p_{0,1}p_{2,3}}. \]
  Then in characteristic $p=3$ we have:
  \[ F_*\OO_X = \OO \oplus \OO(-1)^{44} \oplus \OO(-2)^{20} \oplus A^4 \oplus B^4, \]
  where $A$ and $B$ are rank-2 indecomposable bundles (c.f.~\cite{RSVdB22}).
\end{exa}

@@

  For instance, consider $X = \Bl_4\P^2$, the blowup of $\P^2$ at 4 general points. We will be working over the $\ZZ^5$-graded Cox ring
  \[ S = k[x_1,\dots,x_{10}]/\text{(five quadric Pl\"ucker relations)} \]
  with degrees
    \[
    \left(\!\begin{array}{rrrrrrrrrr}
      0&0&0&0&1&1&1&1&1&1 \\
      1&0&0&0&-1&-1&-1&0&0&0 \\
      0&1&0&0&-1&0&0&-1&-1&0 \\
      0&0&1&0&0&-1&0&-1&0&-1 \\
      0&0&0&1&0&0&-1&0&-1&-1
    \end{array}\!\right).
    \]
    Then in characteristic 2 we have:
    \begin{align*}
      F_*^2\OO_X = {\OO_{X}^{1}}
      &\oplus {\OO_{X}^{2}\ \left(-2,\,1,\,1,\,1,\,1\right)} \oplus {\OO_{X}^{2}\ \left(-1,\,0,\,0,\,0,\,1\right)} \\
      &\oplus {\OO_{X}^{2}\ \left(-1,\,0,\,0,\,1,\,0\right)} \oplus {\OO_{X}^{2}\ \left(-1,\,0,\,1,\,0,\,0\right)} \\
      &\oplus {\OO_{X}^{2}\ \left(-1,\,1,\,0,\,0,\,0\right)} \oplus B \oplus G,
    \end{align*}
    where $B, G$ are rank-3 and rank-2 indecomposable modules, as calculated in \cite{Hara15}.
\end{exa}

\begin{exa}[Frobenius pushforward on cubic surfaces]
  Let $X$ be a smooth cubic surface. Aside from a single exception in characteristic 0, $X$ will be globally $F$-split, so that any $F^e_*\OO_X $ admits $\OO_X$ as a direct summand.
  The other summands of Frobenius pushforwards of $\OO_X$ have yet to be studied, and in particular it is not known whether such rings should have the finite $F$-representation type property.

  The use of our algorithm to compute examples in small $p$ and $e$ suggest the following behavior:
  $$ F_* \OO_X = \OO_X\oplus M, $$
  with $M$ indecomposable, and furthermore $F_*^e M$ remains indecomposable for all $e\geq 0$. In other words, the indecomposable decomposition of $F^e_* \OO_X$ is
  $$ F_*^e \OO_X \cong \OO_X\oplus M\oplus F_* M\oplus\dots\oplus F_*^{e-1}M. $$
  In particular, this suggests $\OO_X$ will fail to have the finite $F$-representation type property.
  In fact, we believe a similar description holds true for quartic del Pezzos.
\end{exa}
